Output Approximation for PHP at Scale:\\ Experience and Analysis

### 2017-10-11 paper.tex
minor improvements

Knowledge about different output variants is leveraged by a number of
tools and analyses for the domain of web applications. Previous work presented
tools for detecting and locating HTML validation errors
\cite{Nguyen:2011:AFH:2190078.2190142}, computing program slices across
server-side and client-page code \cite{Nguyen:2015:CPS:2786805.2786872}, or
easing development and maintenance by extending IDE support for web applications
with code navigation \cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.
Moreover, knowledge of output variants can increase code coverage for
output-oriented testing \cite{Alshahwan2011}.

Despite the various use cases for analyses based on an approximated output
model, static output approximation for PHP web applications using symbolic
execution has so far only been evaluated for small systems that are not
actively maintained. The tools presented by previous work
\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2011:AFH:2190078.2190142}
though are only practical, if for a given system the symbolic execution engine
is scalable, i.e., will also approximate client page output accurately for
larger and more recent systems with acceptable time and space consumption.

@@

\subsection{Symbolic-Execution-based Tools For PHP}\label{sec:existing_tools}
All tools proposed so far and presented in the following leverage symbolically
approximated client page output. Since any single client page output variant can
be analyzed this enables tool support addressing the web application as a whole
in spite of its staged nature, where parts of the web application are
generated dynamically.
Based on the approximated output representation, subsequent developer tools and
analyses have been proposed:
\begin{itemize}
	\item \emph{PhpSync}: Using all variants of HTML client page output, every
	single one can be statically checked for markup validity, i.e., if the web page
	conforms to syntactical specifications for HTML and other client-side
	languages. After the tool detects validation errors, if possible, auto-fixes
	are provided. Otherwise, it traces validation errors back to
	source code responsible for the defect, which then can be refactored manually
	\cite{Nguyen:2011:AFH:2190078.2190142}.

@@

\section{Scalable Output Approximation?} \label{sec:3}
The main objective of this experience report is to investigate whether we can
build a scalable symbolic execution engine as it is the underlying technique
for static output approximation in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}.
Our vision is a practical symbolic execution, i.e., it computes an accurate
approximation for large and modern PHP systems with reasonable time and space consumption.
We approach this goal by re-implementing the symbolic execution semantics  of
\sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928} and extending it with support for additional language
including object-oriented programming and array operations. We aim to support symbolic
execution for a real-world example system of reasonable size that incorporates
the missing language features described in Section \ref{sec:limitations}.
\textsf{WordPress} is a popular open-source Content Management System (CMS)
with around 300.000 lines of code that provides a vivid
plug-in environment and is actively maintained. The full implementation of
\sf{Oak}, the test suite as well as or test corpus (see Section \ref{sec:experiment_setup}) can be found at
\url{http://www.github.com/smba/oak}.

@@

\section{Evaluation} \label{sec:evaluation}
As already stated in the previous Section \ref{sec:experience_report}, we
encountered several problems during the implementation of our symbolic
execution engine for \sf{WordPress}. The main items of critique of the symbolic
execution engine and tools described in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}
are the missing support for widely-used language features, and the evaluation
only using small-scale systems. Although evaluating a tool for a small systems
can be a valuable proof of concept, it does not necessarily cover the question
of whether a tool is practical on a larger-scale. Therefore in the following we
present our methods to evaluate practicality of our symbolic output
approximation. Moreover, we investigate to which extent the limitations encountered
with \sf{WordPress} apply for a wider choice of larger PHP systems.

@@

Second, we ask: \textit{Do system size and usage of modern language features
factors affect the practicality of our symbolic execution engine?} We
selected both a range of large-scale and small-scale modern PHP systems to
better separate and understand the circumstances under which limitations might affect practicality. For the selection of large-scale systems
we borrowed three systems from a case study by Hills et al.
\cite{hills2015variable} addressing the feature usage patterns and frequencies in PHP systems. Some more detailed description of this
case study can be found in Section \ref{sec:related_work}. For our choice of modern small-scale systems we
selected five systems from a list of recent Content Management Systems (CMS) written in PHP \cite{codegeekz}. For both
selections we were bound to select a smaller number of systems than we intended
to since the parser used in our implementation,
\sf{Quercus}\footnote{\sf{Quercus} is an open-source Java implementation of PHP
5 by Caucho Technology. We reused the parser front-end for our implementation.
\sf{Quercus} can be obtained from \url{http://quercus.caucho.com}.}, did not
support a number of language features.
The full list of PHP systems for that we evaluate our symbolic execution engine
along with descriptive code metrics can be found in Table \ref{corpus}.

@@

	\end{subfigure}
	\caption{Evaluation results for output candidates, include expressions and
	call-graph analysis for not-reached output candidates}
\end{figure}

@@

\paragraph{Dynamic features are problematic.}
From out observations with \sf{WordPress} in Section \ref{sec:3} we learn that
dynamic language features, in particular include expressions and dynamic
function invocations containing symbolic values, often cannot be executed
accurately since include targets or function names respectively are resolved
imprecisely.
For our observations we were able to explain most of the missed
output candidates with indirect function invocations. For include expressions we
are indeed able to resolve most reached expressions successfully, yet we do not
distinguish between static and dynamic includes in this case study as this 
would exceed the scope of our paper. However, we believe it is plausible to
assume that imprecisely resolved include expressions can cause a cascade of
missing files and subsequent includes.
Throughout the evaluation we have seen that dynamic features, are commonly used
in modern PHP systems as they serve a genuine purpose in the programming
language.
Nevertheless their usage in situations where they contain assumed symbolic
values restricts symbolic execution and any tool leveraging results obtained
from that.

@@

Aside from being limited by dynamic features, we exercised replacing dynamic
features with respective static counterparts in order to bypass limitations by
manually providing expressions that can be resolved precisely.To use
approaches incorporating symbolic execution with dynamic web applications in
the long run, more concrete input, similar to test oracles, needs
to be provided to decrease the number of assumed symbolic inputs and avoid
dynamic features being confronted with symbolic values. Existing work approaching
this goal \cite{artzi_finding_2008,artzi_finding_2010,DynamicWassermann}
incorporates dynamic symbolic execution, i.e. concrete execution alongside symbolic execution.
Also, if the usage of tools using static output approximation with symbolic
execution is intended, avoiding dynamic features that are prone to be inaccurately executed for
symbolic execution  may be an adjustment strategy to program in an analyzable
way.

\paragraph{Practical tools require practical evaluation.}
We wanted to address the limited informative value of analyses and tools
presented in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140,minamide_static_2005,wassermann2007sound}
where only small PHP systems have been studied. These papers introduce
techniques for static output approximation of PHP web applications
\cite{minamide_static_2005,Nguyen:2014:BCG:2635868.2635928,wang_automating_2012} or propose tools
based on respective approximations
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140,wassermann2007sound,wassermann_static_2008}.
As these approaches address practical use cases, they require a more and further comprehensive evaluation in order to  better understand possible limitations and assess their practicality, in particular for large-scale systems.

>>>>

Knowledge about different output variants is leveraged by a number of
tools and analyses for the domain of web applications. Previous work presented
tools for detecting and locating HTML validation errors
\cite{phpsync}, computing program slices across
server-side and client-page code \cite{Nguyen:2015:CPS:2786805.2786872}, or
easing development and maintenance by extending IDE support for web applications
with code navigation \cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.
Moreover, knowledge of output variants can increase code coverage for
output-oriented testing \cite{Alshahwan2011}.

Despite the various use cases for analyses based on an approximated output
model, static output approximation for PHP web applications using symbolic
execution has so far only been evaluated for small systems that are not
actively maintained. The tools presented by previous work
\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,phpsync}
though are only practical, if for a given system the symbolic execution engine
is scalable, i.e., will also approximate client page output accurately for
larger and more recent systems with acceptable time and space consumption.

@@

\subsection{Symbolic-Execution-based Tools For PHP}\label{sec:existing_tools}
All tools proposed so far and presented in the following leverage symbolically
approximated client page output. Since any single client page output variant can
be analyzed this enables tool support addressing the web application as a whole
in spite of its staged nature, where parts of the web application are
generated dynamically.
Based on the approximated output representation, subsequent developer tools and
analyses have been proposed:
\begin{itemize}
	\item \emph{PhpSync}: Using all variants of HTML client page output, every
	single one can be statically checked for markup validity, i.e., if the web page
	conforms to syntactical specifications for HTML and other client-side
	languages. After the tool detects validation errors, if possible, auto-fixes
	are provided. Otherwise, it traces validation errors back to
	source code responsible for the defect, which then can be refactored manually
	\cite{phpsync}.

@@

\section{Scalable Output Approximation?} \label{sec:3}
The main objective of this experience report is to investigate whether we can
build a scalable symbolic execution engine as it is the underlying technique
for static output approximation in previous work
\cite{phpsync,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}.
Our vision is a practical symbolic execution, i.e., it computes an accurate
approximation for large and modern PHP systems with reasonable time and space consumption.
We approach this goal by re-implementing the symbolic execution semantics  of
\sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928} and extending it with support for additional language
including object-oriented programming and array operations. We aim to support symbolic
execution for a real-world example system of reasonable size that incorporates
the missing language features described in Section \ref{sec:limitations}.
\textsf{WordPress} is a popular open-source Content Management System (CMS)
with around 300.000 lines of code that provides a vivid
plug-in environment and is actively maintained. The full implementation of
\sf{Oak}, the test suite as well as or test corpus (see Section \ref{sec:experiment_setup}) can be found at
\url{http://www.github.com/smba/oak}.

@@

\section{Evaluation} \label{sec:evaluation}
As already stated in the previous Section \ref{sec:experience_report}, we
encountered several problems during the implementation of our symbolic
execution engine for \sf{WordPress}. The main items of critique of the symbolic
execution engine and tools described in previous work
\cite{phpsync,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}
are the missing support for widely-used language features, and the evaluation
only using small-scale systems. Although evaluating a tool for a small systems
can be a valuable proof of concept, it does not necessarily cover the question
of whether a tool is practical on a larger-scale. Therefore in the following we
present our methods to evaluate practicality of our symbolic output
approximation. Moreover, we investigate to which extent the limitations encountered
with \sf{WordPress} apply for a wider choice of larger PHP systems.

@@

Second, we ask: \textit{Do system size and usage of modern language features
factors affect the practicality of our symbolic execution engine?} We
selected both a range of large-scale and small-scale modern PHP systems to
better separate and understand the circumstances under which limitations might affect practicality. For the selection of large-scale systems
we borrowed three systems from a case study by Hills et al.
\cite{hills2015variable} addressing the feature usage patterns and frequencies in PHP systems. Some more detailed description of this
case study can be found in Section \ref{sec:related_work}. Many of the older small-scale systems in our corpus have been studied before in similar contexts. For instance, Nguyen et al.~\cite{phpsync,Nguyen:2014:BCG:2635868.2635928} have selected among others these systems from \sf{sourceforge.net} while considering different sizes and domains as well as little use of object-oriented constructs. For our choice of modern small-scale systems we
selected five systems from a list of recent Content Management Systems (CMS) written in PHP \cite{codegeekz}. For both
selections we were bound to select a smaller number of systems than we intended
to since the parser used in our implementation,
\sf{Quercus}\footnote{\sf{Quercus} is an open-source Java implementation of PHP
5 by Caucho Technology. We reused the parser front-end for our implementation.
\sf{Quercus} can be obtained from \url{http://quercus.caucho.com}.}, did not
support a number of language features.
The full list of PHP systems for that we evaluate our symbolic execution engine
along with descriptive code metrics can be found in Table \ref{corpus}.

@@

	\end{subfigure}
	\caption{Evaluation results for output candidates, include expressions and
	call-graph analysis for not-reached output candidates}
\end{figure}
\vspace{-3mm}

@@

\paragraph{Dynamic features are problematic.}
From our observations with \sf{WordPress} in Section \ref{sec:3} we learn that
dynamic language features, in particular include expressions and dynamic
function invocations containing symbolic values, often cannot be executed
accurately since include targets or function names respectively are resolved
imprecisely.
For our observations we were able to explain most of the missed
output candidates with indirect function invocations. For include expressions we
are indeed able to resolve most reached expressions successfully, yet we do not
distinguish between static and dynamic includes in this case study as this 
would exceed the scope of our paper. However, we believe it is plausible to
assume that imprecisely resolved include expressions can cause a cascade of
missing files and subsequent includes.
Throughout the evaluation we have seen that dynamic features, are commonly used
in modern PHP systems as they serve a genuine purpose in the programming
language.
Nevertheless their usage in situations where they contain assumed symbolic
values restricts symbolic execution and any tool leveraging results obtained
from that.

@@

Aside from being limited by dynamic features, we exercised replacing dynamic
features with respective static counterparts in order to bypass limitations by
manually providing expressions that can be resolved precisely.To use
approaches incorporating symbolic execution with dynamic web applications in
the long run, more concrete input, similar to test oracles, needs
to be provided to decrease the number of assumed symbolic inputs and avoid
dynamic features being confronted with symbolic values. Existing work approaching
this goal \cite{artzi_finding_2008,artzi-tse10,DynamicWassermann}
incorporates dynamic symbolic execution, i.e. concrete execution alongside symbolic execution.
Also, if the usage of tools using static output approximation with symbolic
execution is intended, avoiding dynamic features that are prone to be inaccurately executed for
symbolic execution  may be an adjustment strategy to program in an analyzable
way.

\paragraph{Practical tools require practical evaluation.}
We wanted to address the limited informative value of analyses and tools
presented in previous work
\cite{phpsync,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140,minamide_static_2005,wassermann2007sound}
where only small PHP systems have been studied. These papers introduce
techniques for static output approximation of PHP web applications
\cite{minamide_static_2005,Nguyen:2014:BCG:2635868.2635928,wang-fse12} or propose tools
based on respective approximations
\cite{phpsync,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140,wassermann2007sound,wassermann_static_2008}.
As these approaches address practical use cases, they require a more and further comprehensive evaluation in order to  better understand possible limitations and assess their practicality, in particular for large-scale systems.

### 2017-10-11 related.tex
minor improvements

\paragraph{Symbolic execution}
A literature review on symbolic execution is provided by Cadar in a
recent overview {\em et al.}~\cite{CadarSen2013}, as well as in a more
tool-centered perspective~\cite{Cadar2011}.
%
Symbolic execution \cite{King1976,Darringer1978} remained a
conceptual approach for decades, but has been refined and extended
over the last years. Several approaches incorporating constraint
solving and leveraging concrete execution alongside symbolic execution
emerged and enable automated test case generation; the scalability of
these approaches is still challenged by path explosion and constraint
solving.

\paragraph{Dynamic symbolic execution} Cedar and Sen~\cite{CadarSen2013} incorporates symbolic execution
alongside concrete execution and has shown promising results in
automated test
generation~\cite{artzi_finding_2008,artzi_finding_2010,DynamicWassermann}
for PHP web applications and might help mitigate the conceptual
limitations faced with symbolic values and dynamic features for future
work on output-oriented symbolic execution.

@@

In another approach by Wang et al. \cite{wang_automating_2012}, 
output for a web application is approximated using a hybrid approach: A
dynamic web page is executed with concrete input and the execution is
recorded at run-time. Changes in the client-side output then can be mapped to 
corresponding PHP code using static impact analysis.

Several string taint-analysis techniques were built for PHP web
programs and software-security
problems~\cite{kieyzun09,su-icse08,xieaiken06,tevfik11}.  The key idea
behind string tainting is to check the code expression by expression,
variable by variable to have a list of all the ones potentially
influenced by outside input.  If these variables are used in running a
dangerous command, we will have potentially dangerous tainted
variables.
%
Gould {\em et al.}~\cite{gould04} use string analysis to guarantee
well-typed SQL queries generated by a Java program. The type system
in~\cite{tabuchi03} is based on regular expressions with string
concatenation and pattern matching.

>>>>

\paragraph{Symbolic execution}
A literature review on symbolic execution is provided by Cadar and Sen~\cite{CadarSen2013} in a
recent overview, as well as in a more
tool-centered perspective~\cite{Cadar2011}.
%
Symbolic execution \cite{King1976,Darringer1978} remained a
conceptual approach for decades, but has been refined and extended
over the last years. Several approaches incorporating constraint
solving and leveraging concrete execution alongside symbolic execution
emerged and enable automated test case generation; the scalability of
these approaches is still challenged by path explosion and constraint
solving.

\paragraph{Dynamic symbolic execution} Cadar and Sen~\cite{CadarSen2013} incorporates symbolic execution
alongside concrete execution and has shown promising results in
automated test
generation~\cite{artzi_finding_2008,artzi-tse10,DynamicWassermann}
for PHP web applications and might help mitigate the conceptual
limitations faced with symbolic values and dynamic features for future
work on output-oriented symbolic execution.

@@

In another approach by Wang et al. \cite{wang-fse12}, 
output for a web application is approximated using a hybrid approach: A
dynamic web page is executed with concrete input and the execution is
recorded at run-time. Changes in the client-side output then can be mapped to 
corresponding PHP code using static impact analysis.

Several string taint-analysis techniques were built for PHP web
programs and software-security
problems~\cite{kieyzun09,wassermann_static_2008,xieaiken06,tevfik11}.  The key idea
behind string tainting is to check the code expression by expression,
variable by variable to have a list of all the ones potentially
influenced by outside input.  If these variables are used in running a
dangerous command, we will have potentially dangerous tainted
variables.
%
Gould {\em et al.}~\cite{gould04} use string analysis to guarantee
well-typed SQL queries generated by a Java program. The type system
in~\cite{tabuchi03} is based on regular expressions with string
concatenation and pattern matching.
